---
title: "bootcamp_day2"
author: "Ivan"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
here::here("data","gapminder5.csv")

# here::here replace the unique home that you can distribute across
gapminder <- read.csv(here::here("data",
                                 "gapminder5.csv"))
```

```{r}
str(gapminder)
```

## Converting Factors to Char

```{r}
gapminder$country<-as.character(gapminder$country)
gapminder$continent<-as.character(gapminder$continent)

```

## Create logged variables with a for loop
```{r}
for (i in 1:nrow(gapminder)){
  gapminder[i,'log_gdpPercap']<-log(gapminder[i,'gdpPercap'])
  gapminder[i,'log_pop']<-log(gapminder[i,'pop'])
  
}

```

## Vectorization way of taking log
```{r}
gapminder$vec_log_gdpPercap <- log(gapminder$gdpPercap)
all(gapminder$vec_log_gdpPercap == gapminder$log_gdpPercap)

```

## Find mean life expectancy for each year
```{r}

years <- unique(gapminder$year)

for (i in years) {
  mean_le <- mean(gapminder$lifeExp[gapminder$year == i], 
                  na.rm = T)
  print(paste0(i, ": ", mean_le))
}

```

```{r}
conts <- unique(gapminder$continent)

for (i in conts) {
  mean_le <- mean(gapminder$lifeExp[gapminder$continent == i], 
                  na.rm = T)
  print(paste0(i, ": ", mean_le))
}

```

```{r}
for (i in conts) {
  print(paste0("Continent: ", i))
  for (j in years) {
    mean_le <- mean(gapminder$lifeExp[gapminder$continent == i & 
                                          gapminder$year == j], 
                      na.rm = T)
    print(paste0(j, ": ", mean_le))
  }
  }


```

```{r}
for (i in conts) {
  print(paste0("Continent: ", i))
  for (j in years) {
    mean_le <- sd(gapminder$lifeExp[gapminder$continent == i & 
                                          gapminder$year == j], 
                      na.rm = T)
    print(paste0(j, ": ", mean_le))
  }
  }


```

## apply family only hides loops
```{r}
vars <- gapminder[, c("lifeExp", "pop", "gdpPercap")]
apply(vars, 2, mean,na.rm = T) # apply takes in a matrix, so 
# take whatever and coerce into matrix. So cannot take Char value.

# lapply
lapply(gapminder, mean,na.rm = T) # return a list

# sapply
sapply(years, function(x) mean(gapminder$lifeExp[gapminder$year == x])) # unlist lapply


```

## While loop
```{r}
i <-  1952 # define the interator

while (i < 1987) {
  sd_lf <- sd(gapminder$lifeExp[gapminder$year == i])
  print(paste0(i, ": ", sd_lf)
        )
  i <- i + 5 # increase the iterator by the interval between years
}

```

```{r}
i <-  1987 # define the interator

while (i <= 2002) {
  sd_lf <- sd(gapminder$lifeExp[gapminder$year == i])
  print(paste0(i, ": ", sd_lf)
        )
  i <- i + 5 # increase the iterator by the interval between years
}

```


```{r}
for (i in years) {
  if (i >= 1987) {
    mean_pop <- mean(gapminder$lifeExp[gapminder$year == i])
    print(paste0(i, ": ", mean_pop))
  } else {
    print("Sorry, year is less than 1987")
  }
}


```

## Function
```{r}
get_values <-
  function(df, variable = "continent") {
    vals <- unique(df[[variable]])
    print(paste0(variable, ": ", vals))
  }


```

```{r}

report_mean_sd <- 
  function(df, variable, country) {
    var <- df[[variable]][df$country == country]
    m_le <- mean(var)
    median_le <- median(var)
    min_le <- min(var)
    print(min_le)
    max_le <- max(var)
    
    cat("Country:", country, 
        "\nMean Life Expectancy:", m_le,
        "\nmedian Life Expectancy:", median_le,
        "\nmin Life Expectancy:", min_le,
        "\nmax Life Expectancy:", max_le
        )
  }

report_mean_sd(gapminder, "lifeExp", "Afghanistan")


```

## Exercises

### Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

```{r}
if(any(gapminder$year==2002)){ print('True')}
if(any(gapminder$year==2012)){ print('True')} else {print('False')}

```

###Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset



```{r}
mean_pop<-mean(gapminder$pop)
library(tidyverse)
gapminder %>% filter(pop<mean_pop) %>% group_by(country) %>% summarise(mean=mean(lifeExp))

```
### Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.
```{r}
for (i in conts) {
  mean_le<-mean(gapminder[gapminder$continent==i,'lifeExp'])
  if(mean_le < 50){ 
    print(paste0(i, ':' ,'smaller than 50'))
  }else if (mean_le > 70) {
      print(paste0(i, ':' ,'greater than 70'))
  }else{
      print(paste0(i, ':' ,'between 50 and 70'))
    }
}



```

### Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset. Hint: Use mode() or class() to get the class of the data in each column. Remember that names() or colnames() returns the name of the columns in a dataset.

```{r}

prt_df <- function(df){
  nam<-names(df)
  for (n in nam){
    print(n)
    print(class(df[[n]]))
  }
}

prt_df(gapminder)
```
### Create a function that given a vector will print the mean and the standard deviation of a vector, it will optionally also print the median. Hint: include an argument that takes a boolean (TRUE/FALSE) operator and then include an if statement.

```{r}
prt_sd_mean <- function(v, med = F){
  print(mean(v))
  print(sd(v))
  if(med){
    print(median(v))
  }
}

prt_sd_mean(sample(1:20,10))
prt_sd_mean(sample(1:20,10),med = T)


```
### What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

```{r}
gapminder$vec_log_gdpPercap <- log(gapminder$gdpPercap)
gapminder$vec_log_pop<- log(gapminder$pop)
cor(gapminder$vec_log_gdpPercap,gapminder$vec_log_pop)

viz_lm <-
  function(df, dv, iv, year=0, year_flag = F) {
    if(year_flag){
      dat <- df[df[["year"]] == year, ]
    }else{
      dat<-df
    }
    
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = year,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }

pred <- lm(gapminder$vec_log_gdpPercap~gapminder$vec_log_pop)
summary(pred) # positively related


for (i in years) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i, year_flag = T)
} # Yes, slope changes


```
### Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.
```{r}
gapminder[gapminder$country %in% c('United States','Canada','Mexico'), 'continent']<-c('North America')
gapminder[gapminder$continent=='Americas', 'continent'] <- 'South America'

new_conts <- unique(gapminder$continent)

viz_lm <-
  function(df, dv, iv, continent) {
    
    dat <- df[df[["continent"]] == continent, ]
   
    
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = continent,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }

for (i in new_conts){
  viz_lm(gapminder,"lifeExp", "gdpPercap",i)
}
# Yes it does vary 
```





